---
title: "An Analysis of Fuel Efficiency"
author: "Clyde Tressler"
date: "December 18, 2015"
output: pdf_document
---

##Executive Summary

MotorTrend magazine data from observations of 32 models of cars were analyzed to help understand the characteristics that account for fuel efficiency. Of particular interest is whether there is an effect associated with automatic versus manual transmission. After we account for a strong relationship between vehicle weight and transmission type, we claim there is a statistically significant increase in the mean value of gas mileage attributable to manual transmissions, which we report as **8.03 mpg**, with a **P-value $\leq$ 0.03594**.

###Data Processing and Initial Exploratory Analyses
This report omits some code for brevity. The entire R markdown file can be found [here](https://github.com/amaglamatedclyde/mtcars).

Here we show the first several rows of the dataset.

```{r, include=F}
library(knitr)
library(dplyr)
library(car)
data(mtcars)
attach(mtcars)
```

```{r, message=FALSE, echo=FALSE}
head(mtcars, 3)
```

In the Appendix we present a table explaining the meaning of variable names. Several binary and discrete variables were transformed from numeric vectors to factor variables for this analysis.

```{r, include=FALSE}
mtcars$am <- as.factor(mtcars$am)
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$carb <- as.factor(mtcars$carb)
```

Also in the Appendix we present a box and whisker plot of the two transmission types demonstrating that the median mpg of the automatic group lies below that of the manual group.

```{r, include=F}

am1 <- subset(mtcars, am==1)
am0 <- subset(mtcars, am==0)
t.test(am1$mpg, am0$mpg)
```

```{r, include=F}
diff.means <- round(mean(am1$mpg)-mean(am0$mpg), 2)
```

We report a p-value for a t-test with unequal variances as 0.001374 and therefor reject the null hypothesis at the 95% confidence level, to conclude that the difference between the means of the mileage for the two transmission types, `r diff.means` mpg, is significant. 

This is an important first clue, but, as we will show, it does not present a complete picture.

###Developing a Multivariate Linear Model
Next we turn to multivariate regression to develop a model that will help us understand whether confounding variables are contributing **[omitted variable bias](https://en.wikipedia.org/wiki/Omitted-variable_bias)**. We tabulate the correlation of the numeric variables. Many of these measure closely-related properties, as these values clearly indicate. 

<div align="center">**Correlation Table of mtcars Data**</div>
```{r, echo=FALSE, fig.align="top", fig.height=2}
kable(cor(mtcars[sapply(mtcars, class)=="numeric"]))
```


###Feature Selection
We first consider the apriori exclusion of certain variables. For example, the quarter second mile time is influenced by driver skill and is heavily biased in favor of manual transmission. This is not meaningful to our goal of quantifying transmission effects under normal driving conditions, so we choose to exclude it. 

We also know that 'drat,' the rear axle ratio, is a downstream variable determined by the upstream drive train, including the transmission, so the effect of this variable is captured by other features.

We select weight as our first predictor since it shows the highest correlation with mpg. 

When we examine the data sorted by weight, we see that 9 of the top 10 heaviest cars have automatic transmissions, while 9 of lightest 10 have manual. This leads us to consider the addition to our model of an interaction term between weight and transmission type. In the Appendix we show a visualization of this relationship, where the two transmission types exhibit a distinct separation in a plot of weight vs MPG. 

<div align="center">**Sorting the Data by Vehicle Weight**</div>
```{r Sorting By Weight, echo=F}
kable(mtcars %>% add_rownames(var="rownames") %>% select(rownames, wt, am, hp, mpg) %>% arrange(-wt) %>% head(10))
kable(mtcars %>% add_rownames(var="rownames") %>% select(rownames, wt, am, hp, mpg) %>% arrange(-wt) %>% tail(10))
```

The effect of other variables on the model was examined systematically and led us to choose to include horsepower. 

```{r, echo=F, message=F, include=F}
model2 <- lm(mpg~wt+am+cyl, mtcars)
model3 <- lm(mpg~wt+am+disp, mtcars)
model4 <- lm(mpg~wt+am+hp, mtcars)
# summary(model4)
model5 <- lm(mpg~wt+am+vs, mtcars)
model6 <- lm(mpg~wt+am+gear, mtcars)
model7 <- lm(mpg~wt+am+carb, mtcars)
# lapply(list(model2, model3, model4, model5, model6, model7), function(m) summary(m))
lapply(list(model4, model5), function(m) summary(m)$coef)
```

####Diagnostics

ANOVA testing of nested models was employed to choose which terms are significant in the presence of others. 

```{r, echo=F, message=F, include=F}
fit <- lm(mpg~wt, mtcars)
fit2 <- update(fit, mpg~wt+hp)
fit3 <- update(fit, mpg~wt+hp+am)
fit4 <- update(fit, mpg~wt+hp+am+wt*am)
fit5 <- update(fit, mpg~wt+hp+am+wt*am+hp*am)
anova(fit,fit2,fit3,fit4,fit5)

```
Once we satisfy the required inclusion of transmission type, it was seen that the addition of the interaction term (wt*am) was also needed. 

The reader is again referred to the full R markdown file for the code demonstrating the model search procedure and the ANOVA table.

###High Influence Points
In the Appendix we plot residuals vs hat-values for the model **mpg~wt+hp+am+wt\*am** and identify the Maserati Bora as a high influence point. This is a high-performance vehicle and is the only vehicle in the data set with 8 carburetors. Furthermore, the dfbeta of this point for the coefficient of transmission type, -2.2929, lends motivation to our choice to remove it.

The residuals vs Fitted and Q-Q plots shown in the Appendix indicate the distribution of the residuals is close to normal. 
```{r, echo=F, message=F}
mtcars2 <- mtcars[-31,]
candidate.model <- lm(mpg~wt+hp+am+wt*am, mtcars2)

hp.squared <- lm(mpg~wt+hp+am+wt*am+I(hp*hp), mtcars2)
summary(candidate.model)
# confint(candidate.model)
# summary(hp.squared)
```

###Conclusion
We have shown that a linear model using weight, horsepower, transmission type, and an interaction term between weight and transmission fits observations in the data set well, with an R-squared value of 0.868, indicating 86.8% of the total variance in mpg has been explained. The 95% confidence intervals for the coefficients show no instances where the intervals cross zero and the residuals plots show no discernible patterns. 

To quantify the impact of transmission type on gas mileage using what we've learned from this model, we subset the data by transmission type and examine a linear model with weight and horsepower as predictors for each group separately. Horsepower is no longer a significant predictor for the manual transmission group. We calculate the mean mpg for each model evaluated at mean weight and horsepower. We report a p-value based on the highest value of the p-values for all terms in the calculation. The code is shown in the R markdown file previously cited.

```{r, echo=F, include=F}
am1 <- subset(mtcars2, am==1)
am0 <- subset(mtcars2, am==0)
automatic <- lm(mpg~wt+hp, am0)
manual <- lm(mpg~wt, am1)
mean.data.am0 <- data.frame(mean(am0$wt), mean(am0$hp))
names(mean.data.am0) <- c("wt", "hp")
mean.data.am1 <- data.frame(mean(am1$wt))
names(mean.data.am1) <- c("wt")
am0.bar <- predict.lm(automatic, mean.data.am0)
am1.bar <- predict.lm(manual, mean.data.am1)
confint(automatic)
confint(manual)
```

```{r}
summary(automatic)
summary(manual)
```
Accounting for weight and horsepower, we now report a value of **`r round(am1.bar-am0.bar, 2)`mpg** for the increase in the mean mpg value for manual transmission.

##Appendix

###[Table 1: Variables](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)
Variable Name | Description
------------- | -------------
mpg|Miles/(US) gallon
cyl|Number of cylinders
disp|Displacement (cu.in.)
hp|Gross horsepower
drat|Rear axle ratio
wt|Weight(lb/1000)
qsec|1/4 mile time
vs|Cylinder alignment, V or Straight
am|Transmission (0 = automatic, 1 = manual)
gear|Number of forward gears
carb|Number of carburetors

<div align="center">
##Pairs Plot for mtcars
```{r, echo=F, fig.height=8, fig.width=8}
pairs(mtcars, col = "firebrick")
```

##Box and Whisker plot of MPG vs Transmission Type

```{r, echo=F, fig.width=6,fig.height=6, fig.align='center'}
plot(mtcars$am, mtcars$mpg, col="steelblue", xaxt = "n", main="MPG vs Transmission Type", ylab="MPG")
axis(1, at=1:2, labels=c("Automatic", "Manual"))
```

##Interactions for Weight and Transmission and for Horsepower and Transmission
```{r, echo=F, message=F}
library(lattice)
require(gridExtra)
par(mfrow=c(1,2))
plot(mtcars$am, mtcars$wt, col="steelblue", xaxt = "n", main="Weight vs Transmission Type", ylab="Weight in 1/1000lb")
axis(1, at=1:2, labels=c("Automatic", "Manual"))
plot(mtcars$am, mtcars$hp, col="firebrick", xaxt = "n", main="Horsepower vs Trans. Type", ylab="Horsepower")
axis(1, at=1:2, labels=c("Automatic", "Manual"))
plot1 <- xyplot(mpg~wt,data=mtcars,groups=factor(am,labels=c("A","M")),
       pch=20,auto.key=list(columns=2),type=c("p","g"), cex=2)
plot2 <- xyplot(mpg~hp,data=mtcars,groups=factor(am,labels=c("A","M")),
       pch=20,auto.key=list(columns=2),type=c("p","g"), cex=2)
grid.arrange(plot1,plot2, ncol=2)
```

##Checking for High Influence Points

```{r, echo=F, fig.height=8, fig.width=8}
plot(hatvalues(fit4), fit4$residuals, pch=19, col="firebrick", main = "HatValues vs Residuals")
text( hatvalues(fit4), fit4$residuals, rownames( mtcars ), pos= 3, col="steelblue")
```

##Comparing models with and without Horsepower Squared Term
```{r, echo=F, message=F, fig.height=8}
par(mfrow=c(2,2))
plot(candidate.model)
```

```{r, echo=F, message=F, fig.height=8}
par(mfrow=c(2,2))
plot(hp.squared)
```
</div>